---
title: "Kata: Apply SPC to Datadog SLOs"
output: html_document
date: "2023-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(lubridate)
library(scales)
library(dplyr)
library(glue)
library(httr)
library(purrr)
```

# Iteration 0: Pull SLI history from the API

Use the `httr` library to make an [SLO history request](https://docs.datadoghq.com/api/latest/service-level-objectives/#get-an-slos-history).
This returns data shaped like (only relevant data is shown):

```json
{
  "data": {
    "series": {
      "numerator": {
        "values": [ count, ... ],
      },
      "denominator": {
        "values": [ count, ... ],
      },
      "times": [ timestampInMilliseconds, ... ]
    }
  }
}
```

This iteration show how to make the request. Data frames will be created in the
following iterations.

```{r , include=TRUE}
getSLOHistory <- function(slo = NULL, to = NULL, from = NULL) {
  apiKey <- Sys.getenv("DATADOG_API_KEY")
  applicationKey <- Sys.getenv("DATADOG_APP_KEY")

  if (apiKey == "" || applicationKey == "") {
    stop("DATADOG_API_KEY or DATADOG_APP_KEY missing")
  }

  r <-
    GET(
      glue("https://api.datadoghq.com/api/v1/slo/{slo}/history"),
      query = list(from_ts = as.integer(from),
                   to_ts = as.integer(to)),
      add_headers("DD-API-KEY" = apiKey, "DD-APPLICATION_KEY" = applicationKey)
    )

  stop_for_status(r)

  json <- content(r)
}

json <- getSLOHistory(
  slo = "5439926203625d579e0f33d44afcad49",
  from = now() - days(60),
  to = now()
)

summary(json)
```

# Iteration 1: Parse into a Dataframe

The useful data is in `data.series.times`, `data.series.numerator.values`,
and `data.series.denominator.values`. Each is a list. The three may be combined
into a data frame. Then the `SLI` value is calculated.

This iteration uses the `purrr` for iteration and manipulation.

```{r, iteration1, include = TRUE}
timestamps <- json %>%
  pluck("data", "series", "times") %>%
  map(\(x) as_date(as_datetime(x / 1000))) %>%
  list_c

numerators <- json %>%
  pluck("data", "series", "numerator", "values") %>%
  list_c

denominators <- json %>%
  pluck("data", "series", "denominator", "values") %>%
  list_c

history <- data.frame(
  Date = timestamps,
  Numerator = numerators,
  Denominator = denominators
) %>%
  mutate(SLI = Numerator / Denominator)
```

# Iteration 2: Graph the SLI

Create a simple line plot with the Date on the X axis and the
SLI (as percentage) on the Y axis. Add a blue line for the SLO. How many points
meet the SLO?

```{r, iteration2, include = TRUE}
ggplot(history, aes(x = Date, y = SLI)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.9999, linetype = "solid", color = "blue") +
  scale_y_continuous(labels = label_percent()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

# Iteration 3: Create Control Chart

See the other SPC katas for example. This code calculates the upper and lower
natural process limits using SPC constants. They are added to the chart in red,
along with the mean in black.

This demonstrates what we can expect from this process in it's current
condition.

```{r iteration3, include = TRUE}
history <- history %>%
  mutate(
    UNPL = mean(SLI) + 2.66 * mean(abs(diff(SLI))),
    LNPL = mean(SLI) - 2.66 * mean(abs(diff(SLI))),
    Mean = mean(SLI)
  )

ggplot(history, aes(x = Date, y = SLI)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = label_percent()) +
  geom_hline(aes(yintercept = UNPL), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = LNPL), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = Mean), linetype = "solid") +
  geom_hline(yintercept = 0.9999, linetype = "solid", color = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

How far is the blue line from the control limits (red dashed lines)?

# Iteration 4: Color Out of Control Points

Add a column for points outside the control limits. Use that as an factor to
manually color points red (out of control) and black (under control).

```{r iteration3, include = TRUE}
history <- history %>%
  mutate(Violation = SLI > UNPL | SLI < LNPL)

ggplot(history, aes(x = Date, y = SLI)) +
  geom_line() +
  geom_point(aes(color = factor(Violation)), show.legend = FALSE) +
  geom_hline(aes(yintercept = UNPL), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = LNPL), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = Mean), linetype = "solid") +
  geom_hline(yintercept = 0.9999, linetype = "solid", color = "blue") +
  scale_y_continuous(labels = label_percent()) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

# Iteration 5: Create a Weekly Rollup

Datadog coaxes you to look at 7, 30, and 90 day SLOs. The API provides daily
values. Use the `purrr` to group points by the week, then calculate the SLI for
each week. `foor_date` calculates "start of the week on Monday" for each date.

```{r iteration5, include = TRUE}
weeklyHistory <- history %>%
  group_by(Week = floor_date(Date, unit = "week", week_start = 1)) %>%
  summarise(SLI = sum(Numerator) / sum(Denominator)) %>%
  mutate(
    UNPL = mean(SLI) + 2.66 * mean(abs(diff(SLI))),
    LNPL = mean(SLI) - 2.66 * mean(abs(diff(SLI))),
    Mean = mean(SLI)
  ) %>%
  mutate(Violation = SLI > UNPL | SLI < LNPL)

ggplot(weeklyHistory, aes(x = Week, y = SLI)) +
  geom_line() +
  geom_point(aes(color = factor(Violation)), show.legend = FALSE) +
  geom_hline(aes(yintercept = UNPL), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = LNPL), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = Mean), linetype = "solid") +
  geom_hline(yintercept = 0.9999, linetype = "solid", color = "blue") +
  scale_y_continuous(labels = label_percent()) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

# Iteration 6: Create a Heads-up-Display of Mulitple SLOs

Now scale the approach to multiple SLOs. The work in previous iterations is
refactored into a single function. The function takes an SLO id and timestamps,
then returns a weekly rollup.

These may be combined into a single dataframe then "split graph" using
`facet_wrap`. This creates a nice view of many SLOs at once.

This time, add a blue line for the SLO itself. How many points meet the SLO?

```{r iteration5, include = TRUE}
getSLOWeeklyHistory <- function(slo = NULL, to = NULL, from = NULL) {
  apiKey <- Sys.getenv("DATADOG_API_KEY")
  applicationKey <- Sys.getenv("DATADOG_APP_KEY")

  if(apiKey == "" || applicationKey == "") {
    stop("DATADOG_API_KEY or DATADOG_APP_KEY missing")
  }

  r <-
    GET(
      glue("https://api.datadoghq.com/api/v1/slo/{slo}/history"),
      query = list(from_ts = as.integer(from),
                   to_ts = as.integer(to)),
      add_headers("DD-API-KEY" = apiKey, "DD-APPLICATION_KEY" = applicationKey)
    )

  stop_for_status(r)

  json <- content(r)

  timestamps <- json %>%
    pluck("data", "series", "times") %>%
    map(\(x) as_date(as_datetime(x / 1000))) %>%
    list_c

  numerators <- json %>%
    pluck("data", "series", "numerator", "values") %>%
    list_c

  denominators <- json %>%
    pluck("data", "series", "denominator", "values") %>%
    list_c

  history <- data.frame(
      Date = timestamps,
      Numerator = numerators,
      Denominator = denominators
    ) %>%
    mutate(SLI = Numerator / Denominator) %>%
    group_by(Week = floor_date(Date, unit = "week", week_start = 1)) %>%
    summarise(SLI = sum(Numerator) / sum(Denominator)) %>%
    mutate(
      UNPL = mean(SLI) + 2.66 * mean(abs(diff(SLI))),
      LNPL = mean(SLI) - 2.66 * mean(abs(diff(SLI))),
      Mean = mean(SLI)
    ) %>%
    mutate(Violation = SLI > UNPL | SLI < LNPL)
}

# This repeated structured could be refactored into some sort of Key/Value map
# connected to a map and reduction function. Something for a separate kata.
serviceA <- getSLOWeeklyHistory(
  slo = "5439926203625d579e0f33d44afcad49",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service A"
  )

serviceB <- getSLOWeeklyHistory(
  slo = "bc5d0ee9a9955dcb8b5c9a53ec2ac8c3",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service B"
  )

serviceC <- getSLOWeeklyHistory(
  slo = "3351084b596b5949ad81b719b4d00afe",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service C"
  )

serviceD <- getSLOWeeklyHistory(
  slo = "da5550ca13bc5b1cb13d383ab8fd4b6d",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service D"
  )

serviceE <- getSLOWeeklyHistory(
  slo = "7bc3e8a86b4255619d4ae3303a196717",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service E"
  )

serviceF <- getSLOWeeklyHistory(
  slo = "1b97b3eaf254528c8fc0455f81b706c8",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service F"
  )

serviceG <- getSLOWeeklyHistory(
  slo = "f68436d700275ff0b4453892146878df",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service G"
  )

serviceH <- getSLOWeeklyHistory(
  slo = "5855af30c4595371a01c009bb7c5af6e",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service H"
  )

serviceI <- getSLOWeeklyHistory(
  slo = "ee92ccb93a61518ebdd3e57b4b5a6cab",
  from = now() - days(90),
  to = now()
  ) %>% mutate(
    Service = "Service I"
  )

combinedWeekly <- rbind(serviceA, serviceB, serviceC, serviceD, serviceE, serviceF, serviceG, serviceH, serviceI)

ggplot(combinedWeekly, aes(x = Week, y = SLI)) +
  geom_line() +
  geom_point(aes(color = factor(Violation)), show.legend = FALSE) +
  geom_hline(aes(yintercept = UNPL), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = LNPL), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = Mean), linetype = "solid") +
  geom_hline(yintercept = 0.9999, linetype = "solid", color = "blue") +
  scale_y_continuous(labels = label_percent()) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  facet_wrap(vars(Service), scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
