---
title: "Kata: Tidy a Spreadsheet for Plotting"
output: html_document
date: "2023-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(lubridate)
library(scales)
library(dplyr)
library(tidyr)
library(readr)
library(glue)
```

# Background

This kata covers common tasks for working columnar reports provided as CSVs.
Columns are typically added for dates and rows for various items. Data may be
missing, malformed, or parsed incorrectly. There may be entire rows or columns
with empty data.

The data represents usage per project over the last year. Produce a stacked bar
chart for the top N projects.

# Increment 1: Parsing the Data

## Iteration 0: Reading CSV

Use the build in `read.csv` function. This function returns a Data frame. That's
the base for further manipulations.

```{r iteration0, include = TRUE}
usage <- read.csv('circleci_usage.csv')

summary(usage)
```

This creates a few challenges:

1. Numbers are parsed as strings (they contain `,` seperators)
2. Header columns are named strangely (some now include `X`)
3. There are `NA` columns (seemingly dynamically created by the parser)
4. There is a useless second row (seemingly used as a header)

The next steps is to clean the data needed to pivot the table.

## Iteration 1: Remove Junk Rows & Columns

Enter `dplyr` for manipulating the data frame. `filter()` removes the header
row using `row_number()`. `rename` will fix the first missing column name.
`select()` limits data to a set of
columns. That will remove the known junk `Credit.Usage.Activity.Dt.Month`
column. `select()` may also be used with the `matches()` helper for regex.
The second `select()` call removes all the empty columns on the right.

```{r iteration1, include = TRUE}
usage1 <- usage %>%
  filter(row_number() != 1) %>%
  rename(Project = X) %>%
  select(-Credit.Usage.Activity.Dt.Month) %>%
  select(-matches("X\\.\\d+$"))

summary(usage1)
```

## Iteration 2: Pivot the Table

Change the date columns into row entry. Enter `tidyr`.

```{r iteration2, include = TRUE}
usage2 <- usage1 %>%
  pivot_longer(!Project, names_to = "Date", values_to = "Usage")

summary(usage2)
```

## Iteration 3: Parse Dates

The current data has a `Date` column with values like `X2023.05`. Enter `tidyr`.
`tidyr` includes functions for splitting this data into component parts. Then,
the components may be used to create a date.

`seperate_wider_position()` can tokenize a column into multiple columns based
on position. The first character (`X`) is ignored (not given a mapping). The
next four are assigned to `Year`. The next character (`.`) is ignored. The
next two characters are assigned to `Month`.

`mutate()` adds a new column by parsing the two with `lubridate`. The temporary
`Year` and `Month` columns are discarded with `select()`

```{r iteration3, include = TRUE}
usage3 <- usage2 %>%
  separate_wider_position(Date, c(1, Year = 4, 1, Month = 2)) %>%
  mutate(Date = ym(glue("{Year}.{Month}"))) %>%
  select(c(Project, Date, Usage))

summary(usage3)
```

## Iteration 4: Parse Numbers

The `Usage` columns contains human readable strings and empty strings for zero.
Enter `readr`. It includes `parse_number()` which works perfectly for this case.
Empty values become `NA`. `tidyr` includes the `replace_na` function.

```{r iteration4, include = TRUE}
usage4 <- usage3 %>%
  mutate(Usage = parse_number(Usage)) %>%
  replace_na(list(Usage = 0))

summary(usage4)
```

Tada! Finally, a data set fit for plotting!

# Increment 2: The Bar Chart

## Iteration 5: A stacked bar chart

The simple approach is just creating the bar chart per project.

```{r iteration5, include = TRUE}
ggplot(usage4, aes(x = Date, y = Usage, fill = Project)) +
  geom_bar(stat = "identity")
```

Unfortunately, this is completely unusable due to the total number of projects.
Running this R studio produces a confusing chart. It appears to only display
the legend. Open the output in a new window, then make the window bigger. Then,
the full graph is visible.

Something must be done to limit the number of projects.

## Iteration 6: Using the Top 5

The `group_by()` and `rank()` functions work here. `group_by()` for grouping
each `Date`, then `rank()` descending on `Usage`. Then use `filter` to select
the number of items to show.

```{r iteration6, include = TRUE}
usage6 <- usage4 %>%
  group_by(Date) %>%
  mutate(Rank = rank(-Usage, ties.method = "min")) %>%
  filter(Rank <= 5)

ggplot(usage6, aes(x = Date, y = Usage, fill = Project)) +
  geom_bar(stat = "identity")
```

## Iteration 7: Labeling the Y Axis

The `ggplot` defaults to scientific notation for such large values. The `scales`
package includes useful functions for generating labels. The `label_comma()`
function generates human readable numbers with a comma for us Americans.

```{r iteration7, include = TRUE}
ggplot(usage6, aes(x = Date, y = Usage, fill = Project)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_comma())
```

## Iteration 8: A Porportional Chart

The stacked bar char is great to see the totals. It's useful to view the
relative proportion of each project. Luckily, `geom_bar` can do this
out-of-the-box with `position = "fill"`. `geom_bar` will calculate the
proportion and use a 0-1 y-axis. The `scales` package includes a function for
labeling percentages, so use that for the y axis.

```{r iteration8, include = TRUE}
ggplot(usage6, aes(x = Date, y = Usage, fill = Project)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = label_percent())
```

## Iteration 9: Prettify

The default visuals are OK, though difficult to read with this data set. Add an
outline to each bar and change the color. Setting `color` on `geom_bar` adds a
the outline. That improves visibility of each project inside the bar. Second,
use a different color palette for the `fill`. There are many available.
I chose Brewer with `Set3` because of the number of factors. Not all palettes
have the same number of colors.

```{r iteration9, include = TRUE}
ggplot(usage6, aes(x = Date, y = Usage, fill = Project)) +
  geom_bar(stat = "identity", position = "fill", color = "black") +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "Set3")
```

Now, much better.

## Iteration 10: Grouping Lower Ranks

There is one last thing to do. There are 40 plus Projects in the dataset.
Limiting the chart to the top 5 is mandatory though it excludes the rest. This
is an inaccurate presentation of the data. A better solution is lump the lower
ranks into an "other" category. This way the total usage is accurate without
the added noise from tens of small projects.

Creating the "other" project starts of like the Ranking iteration. Then,
any row with a `Rank` higher than 5 has its `Project` set to `other`. Next, the
data is grouped by `Date` and `Project`. Groups are summarized to the sum of
`Usage`.

Practically speaking, Project ranks 1-5 will have a group with size 1. The
`summary` function here is virtually a no-op. It only matters on lower ranks
that will have multiple rows in the group.

Done!

```{r iteration10, include = TRUE}
usage10 <- usage4 %>%
  group_by(Date) %>%
  mutate(Rank = rank(-Usage, ties.method = "min")) %>%
  mutate(Project = ifelse(Rank > 5, "other", Project)) %>%
  group_by(Date, Project) %>%
  summarize(Usage = sum(Usage), .groups = "drop")

ggplot(usage10, aes(x = Date, y = Usage, fill = Project)) +
  geom_bar(stat = "identity", position = "fill", color = "black") +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "Set3")
```
