---
title: "Kata: Produce a Nicely Colored Control Chart"
output: html_document
date: "2023-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tibble)
library(ggplot2)
library(lubridate)
library(scales)
library(dplyr)

# Data taken from a chapter in "Understanding Statistical Process Control"
# by Donald Wheeler. It has been modified to create out-of-control points.

freight <- tribble(
  ~Month, ~TotalShipments,
  1, 4266,
  2, 4440,
  3, 4896,
  4, 4019,
  5, 4101,
  6, 3755,
  7, 5068,
  8, 4040,
  9, 4038,
  10, 3275,
  11, 4059,
  12, 503,
  13, 5311,
  14, 3977,
  15, 4050,
  16, 4546,
  17, 4658,
  18, 4357,
  19, 9585,
  20, 4196,
)
```

# Coloring Points

Add columns for the Upper & Lower Natural Process Limits to the data frame.
Then, add a `Violation` column. The value is determined by the natural process
limit columns. If the point is out-of-control (i.e. violationg the natural
process limits), then they will be colored red. Other points will be black.

The `Violation` column may be used as an R factor. Explicit color values for the
factor (`TRUE` and `FALSE`) are set with `scale_color_manual`.

```{r coloredPoints, include=TRUE}
freight <- freight %>%
  mutate(
    UNPL = mean(TotalShipments) + 2.66 * mean(abs(diff(freight$TotalShipments))),
    LNPL = mean(TotalShipments) - 2.66 * mean(abs(diff(freight$TotalShipments))),
    Mean = mean(TotalShipments)
  ) %>%
  mutate(Violation = TotalShipments > UNPL | TotalShipments < LNPL)

ggplot(freight, aes(x = Month, y = TotalShipments)) +
  geom_line() +
  geom_point(aes(color = factor(Violation)), show.legend = FALSE) +
  geom_hline(aes(yintercept = UNPL), linetype = "dashed") +
  geom_hline(aes(yintercept = LNPL), linetype = "dashed") +
  geom_hline(aes(yintercept = Mean), linetype = "solid") +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  scale_x_continuous(minor_breaks = NULL, breaks = freight$Month) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red"))
```

# Adding a Ribbon

Control charts typicall use dashed lines for the upper and lower control limits.
This kata replaces that with a ribbon (a horizontal band across the chart). This
may be more visualy appealing.

`geom_ribbon` replaces `geom_hline` in the previous example.

```{r withRibbon, include=TRUE}
ggplot(freight, aes(x = Month, y = TotalShipments)) +
  geom_line() +
  geom_ribbon(aes(ymin = LNPL, ymax = UNPL), fill = "blue", alpha = 0.2) +
  geom_point(aes(color = factor(Violation)), show.legend = FALSE) +
  geom_hline(aes(yintercept = Mean), linetype = "solid") +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  scale_x_continuous(minor_breaks = NULL, breaks = freight$Month) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red"))
```

# Increasing the Size of Out-of-Control Points

Using different colors for points out-of-control is great. This may be enhanced
by dynamically setting the size according to the factor.

Here `scale_size_manual` makes out-of-control points twice as large.

```{r withSize, include=TRUE}
ggplot(freight, aes(x = Month, y = TotalShipments)) +
  geom_line() +
  geom_ribbon(aes(ymin = LNPL, ymax = UNPL), fill = "blue", alpha = 0.2) +
  geom_point(aes(color = factor(Violation), size = factor(Violation)), show.legend = FALSE) +
  geom_hline(aes(yintercept = Mean), linetype = "solid") +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  scale_x_continuous(minor_breaks = NULL, breaks = freight$Month) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  scale_size_manual(values = c("FALSE" = 2, "TRUE" = 4))
```

# Changing Shape of Out-of-Control Points

We can also change the size, color, and shape of the point. Here
`scale_shape_manual` turns the out-of-control points into `X` shapes and others
as `o` shapes.

Here's a link to [the shape
codes](https://blog.albertkuo.me/post/point-shape-options-in-ggplot/)

```{r withShape, include=TRUE}
ggplot(freight, aes(x = Month, y = TotalShipments)) +
  geom_line() +
  geom_ribbon(aes(ymin = LNPL, ymax = UNPL), fill = "blue", alpha = 0.2) +
  geom_point(aes(color = factor(Violation), size = factor(Violation), shape = factor(Violation)), show.legend = FALSE) +
  geom_hline(aes(yintercept = Mean), linetype = "solid") +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  scale_x_continuous(minor_breaks = NULL, breaks = freight$Month) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  scale_size_manual(values = c("FALSE" = 2, "TRUE" = 6)) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 4))
```
